---
title: "Dietary Risk Factor Dose-Response Curves V2.1"
author: "A. Jacob"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

#Step 1 - Import Data
This is an excel version of Table 2 from the manuscript located in the folder "data". 

```{r summary}
associations <- read_excel(file.path("data","Table2_159.xlsx"))
```

#Step 2 - Clean the Data
This step removes food category headers, simplifies the linearity and grade coding, and assigns each risk-outcome pair a numeric code. 

```{r dataclean}
associations <- associations[!is.na(associations$Exposure),]

associations$Linearity <- sub(" .*", "", associations$Linearity)

associations$Linearity <- gsub("[–—−]", "-", 
                            gsub("[[:space:]]+", "", 
                              gsub("\u00A0", "", 
                                trimws(tolower(associations$Linearity)))))  

associations$Grade <- sub("\u00A0.*", "", associations$Grade) #strange spacing in the column
associations$Grade <- sub(" .*", "", associations$Grade)
associations$Grade <- substr(associations$Grade, 1, nchar(associations$Grade) - 1) #remove subscript

associations$code <- row.names(associations)

main_key <- associations[,c(2,3,4,5,6,11,12,15)] #Exposure, Outcome, Dose, Observable Range, Risk Estimate, Grade, Linearity, code
```

#Step 3 - Functions of the Lines
For linear function, the slope is applied from zero to the maximum observable range if provided. RR estimates above the observable range are kept constant with the last known point. 

If no observable range was available, the curve begins at zero and continues infinitely. 

```{r linefunction}
#linear function with estimate only
linear_function <- function(R, d, x, min_dose, max_dose) { #R is relative risk, d is increment, x is intake, min_dose is lowest observable value, max_dose is highest observable value
  #Slope per unit
  beta <- log(R)/d
  #Result vector
  result <- numeric(length(x))
  #If either min or max dose is NA, ignore bounds and use original formula
  if (is.na(min_dose) | is.na(max_dose)) {
    result <- exp(beta * x)
  } else {
    for (i in seq_along(x)) {
      if (x[i] > max_dose) {
        #Above observed range → same as maximum
        result[i] <- exp(beta * max_dose)
      } else {
        #Within observed range → linear slope
        result[i] <- exp(beta * x[i])
      }
    }
  }
  
  return(result)
}

#linear function with estimate and 95% confidence interval
linear_function_CI <- function(R, lower, upper, d, x, max_dose) {
  #Slope and standard error (SE) per unit dose
  beta <- log(R) / d
  se_inc <- (log(upper) - log(lower)) / (2 * 1.96)  # assumes 95% CI
  se_unit <- se_inc / d
  
  #Confidence multiplier
  z <- abs(qnorm((1 - 0.95) /2)) #95% cI
  
  #Apply max_dose if provided
  if (!is.na(max_dose)) {
    dose_eff <- pmin(x, max_dose) #takes the smaller of the two values
  } else {
    dose_eff <- x
  }
  
  #Ln-scale RR and CI
  logRR <- beta * dose_eff
  logRR_lower <- logRR - z * se_unit * dose_eff
  logRR_upper <- logRR + z * se_unit * dose_eff
  
  #Back-transform to RR scale
  data.frame(
    dose = x,
    RR = exp(logRR),
    CI_lower = exp(logRR_lower),
    CI_upper = exp(logRR_upper)
  )
}

#non-linear function - no uncertainty extracted
non_linear_function <- function(x_coords, y_coords) {
  function(z) {
    idx <- findInterval(z, x_coords, rightmost.closed = TRUE)
    ifelse(idx == 0, y_coords[1], 
           ifelse(idx > length(y_coords), y_coords[length(y_coords)], y_coords[idx])) #last point is continuous
  }
} 

```

#Step 4 - Linear Data
All associations are included - either linear or summary estimates provided. 

```{r lineardata}

linear_assoc <- main_key

linear_assoc$`Additional Increment` <- gsub("\\s+", "", linear_assoc$`Additional Increment`) #removes spaces

linear_assoc$`Additional Increment` <- sub("([a-zA-Zµ]+)[0-9]+$", "\\1", linear_assoc$`Additional Increment`) #remove trailing digits (superscript)

linear_assoc <- linear_assoc %>%
  extract(`Additional Increment`, into = c("Dose", "Metric"),
          regex = "([0-9\\.]+)([a-zA-Zµ]+)", remove = TRUE) #dose and metric separated

linear_assoc$Dose <- as.numeric(linear_assoc$Dose)

#note that association code 65 is 100 to 200g

#Manage observable range
linear_assoc <- linear_assoc %>%
  mutate(
    `Observable Range` = str_remove_all(`Observable Range`, ","),
    `Observable Range` = str_trim(`Observable Range`), 
    `Observable Range` = na_if(`Observable Range`, "Not provided"), 
    `Observable Range` = na_if(`Observable Range`, "unclear")
  ) %>%
  extract(
    `Observable Range`, 
    into = c("lower_exp", "upper_exp"), 
     regex = "([0-9]+\\.?[0-9]*)\\s*-\\s*([0-9]+\\.?[0-9]*)",
    convert = TRUE
  )

colnames(linear_assoc)[7] <- "RR" 

linear_assoc <- linear_assoc %>%
  extract(RR, 
          into = c("Estimate", "Lower", "Upper"),
          regex = "([0-9\\.]+)\\s*\\[([0-9\\.]+)-([0-9\\.]+)\\]",
          convert = TRUE)
```

#Step 4B - Linear Functions
Creates a list of the linear function according to their code. Uncertainty is replicated in upper and lower bounds. 

```{r linearfunctions}
linear_functions <- list()

#add observable ranges
for(i in seq_len(nrow(linear_assoc))){
  code_p <- linear_assoc$code[i]
  rr_p <- linear_assoc$Estimate[i]
  dose_p <- linear_assoc$Dose[i]
  min_dos_p <- linear_assoc$lower_exp[i]
  max_dos_p <- linear_assoc$upper_exp[i]
  
  linear_functions[[paste0("function_",code_p)]] <- local({
    R <- rr_p
    d <- dose_p
    min_dose <- min_dos_p
    max_dose <- max_dos_p
    function(x) linear_function(R,d,x,min_dose,max_dose)
  })
}
```

```{r linearfunctions CI}
linear_functions_CI <- list()

for (i in seq_len(nrow(linear_assoc))) {
  code_p <- linear_assoc$code[i]
  rr_p <- linear_assoc$Estimate[i]
  lower_p <- linear_assoc$Lower[i]
  upper_p <- linear_assoc$Upper[i]
  dose_p <- linear_assoc$Dose[i]
  max_dose_p <- linear_assoc$upper_exp[i]
  
  linear_functions_CI[[paste0("function_CI_", code_p)]] <- local({
    R <- rr_p
    lower <- lower_p
    upper <- upper_p
    d <- dose_p
    max_dose <- max_dose_p
    function(x) linear_function_CI(R, lower, upper, d, x, max_dose)
  })
}
```

#Step 5 - Non-Linear Curve Data

```{r nonlinear data}
non_linear_assoc <- main_key[main_key$Linearity=="non-linear",]

non_linear_assoc <- non_linear_assoc %>%
  extract(`Additional Increment`, into = c("Dose", "Metric"),
          regex = "([0-9\\.]+)([a-zA-Zµ]+)", remove = TRUE) #dose and metric separated
```

#Step 5B - Non-Linear Curves

```{r nonlinear curves}
csv_files <- list.files(path = file.path("data", "CSV_coordinates_159"), pattern = "\\.csv$", full.names = TRUE)
csv_list <- lapply(csv_files, read.csv)
names(csv_list) <- tools::file_path_sans_ext(basename(csv_files))
```

#Step 5C - Non-Linear Functions

```{r nonlinear functions}
nonlin_functions <- lapply(csv_list, function(df) {
  x <- df[[1]]  
  y <- df[[2]]  
  non_linear_function(x, y)
})

new_names <- paste0("function_", sub("^code", "", names(csv_list)))
names(nonlin_functions) <- new_names
```

#Step 6 - Plots 

```{r linear}
output_dir <- file.path("C:/Users/aeja/Metrix3_C/plots_database")
dir.create(output_dir, showWarnings = FALSE)

linear_plots <- linear_assoc[linear_assoc$Linearity == "linear",]

for (i in seq_len(nrow(linear_plots))) {
  
  # Row-specific values
  code_p <- linear_plots$code[i]
  R      <- linear_plots$Estimate[i]
  lower  <- linear_plots$Lower[i]
  upper  <- linear_plots$Upper[i]
  d      <- linear_plots$Dose[i]        # increment used in the linear model
  
  unit <- linear_plots$Metric[i]
  if (is.na(unit) || unit == "") unit <- "unknown unit"
  
  exposure <- linear_plots$Exposure[i]
  outcome  <- linear_plots$Outcome[i]  
  title    <- paste(exposure, "and", outcome)
  x_label  <- paste0("Intake ", unit, "/day")
  
  # X axis goes from zero → upper plotting limit equals the increment d
  x_vals <- seq(0, d, length.out = 200)
  
  # Compute estimate + CI using the new function
  df <- linear_function_CI(
    R = R,
    lower = lower,
    upper = upper,
    d = d,
    x = x_vals,
    max_dose = d       # ensures the curve does not extend beyond the increment
  )
  
  # Output file
  png_filename <- file.path(output_dir, paste0("plot_linear_", code_p, ".png"))
  png(png_filename, width = 800, height = 600)
  
  # Plot setup
  plot(
    df$dose, df$RR, type = "n",
    xlab = x_label, 
    ylab = "Relative Risk",
    main = title,
    xlim = c(0, d),
    ylim = range(c(df$CI_lower, df$CI_upper))
  )
  
  # Uncertainty band
  polygon(
    c(df$dose, rev(df$dose)),
    c(df$CI_upper, rev(df$CI_lower)),
    col = rgb(0, 0, 1, 0.2), 
    border = NA
  )
  
  # Main line + points
  lines(df$dose, df$RR, col = "blue", lwd = 2)
  points(df$dose, df$RR, col = "blue", pch = 16, cex = 0.6)
  
  dev.off()
}
```

```{r nonlinear}
#Match CSV List to Non Linear Associations

non_linear_assoc$code_name <- paste0("code",non_linear_assoc$code)

for (code_name in intersect(names(csv_list), non_linear_assoc$code_name)) {
  
  #Data
  csv_data <- csv_list[[code_name]]
  plot_df <- data.frame(
    x_vals <- csv_data[[1]],
    y_vals <- csv_data[[2]]
  )
  #Match 
  meta_row <- non_linear_assoc[non_linear_assoc$code_name == code_name, ]
  if (nrow(meta_row) == 0) next  #Skips if code is not found
  
  unit <- meta_row$Metric
  if (is.na(unit) || unit == "") unit <- "unknown unit"
  
  exposure <- meta_row$Exposure
  outcome <- meta_row$Outcome
  title <- paste0(exposure, " and ", outcome)
  x_label <- paste0("Intake ", unit, "/day")
  
  #Prep plot data
  ord <- order(x_vals)
  x_vals <- x_vals[ord]
  y_vals <- y_vals[ord]
  
  #File name
  png_filename <- file.path(output_dir, paste0("plot_", code_name, ".png"))
  
  # --- Plot creation ---
  png(png_filename, width = 600, height = 400)
  
  plot(
    NULL,
    xlim = c(min(x_vals) - 1, max(x_vals) + 1),
    ylim = range(y_vals) + c(-0.5, 0.5),
    xlab = x_label,
    ylab = "Relative Risk",
    main = title,
    type = "n"
  )
  
  # Draw line segments and points
  for (j in 1:(length(x_vals) - 1)) {
    segments(x_vals[j], y_vals[j], x_vals[j + 1], y_vals[j], col = "blue", lwd = 2)
    points(x_vals[j], y_vals[j], pch = 16, col = "blue")
    points(x_vals[j + 1], y_vals[j], pch = 1, col = "blue")
  }
  
  # Extend last segment
  segments(
    x_vals[length(x_vals)], y_vals[length(y_vals)],
    max(x_vals) + 2, y_vals[length(y_vals)],
    col = "blue", lwd = 2
  )
  points(x_vals[length(x_vals)], y_vals[length(y_vals)], pch = 16, col = "blue")
  
  dev.off()
}

```

#Step 7 - Use and Example

There are now 101 functions that can be used as well as plot outputs for visualisation of curves. Note that uncertainty is not accounted for here and the user will need to determine how they would like to estimate uncertainty in their given study. To use the functions, one simply needs to call it as seen below.

```{r example of function use}
#I would like to know the risk estimate for dementia based on fish intake (this is a linear association - code91)
linear_functions$function_91(10) #10g/day
linear_functions$function_91(50) #50g/day
linear_functions$function_91(100) #100g/day
linear_functions_CI$function_CI_91(10) #gives the estimate along with CI

#What about SSBs intake and Prostate Cancer incidence? (non-linear association - code18)
nonlin_functions$function_18(10)
nonlin_functions$function_18(50)
nonlin_functions$function_18(100)
```
