---
title: "Dietary Risk Factor Dose-Response Curves V2.0"
author: "A. Jacob"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

#Step 1 - Import Data
This is an excel version of Table 2 from the manuscript located in the folder "data". 

```{r summary}
associations <- read_excel(file.path("data","Table2_all.xlsx"))
```

#Step 2 - Clean the Data
This step removes food category headers, simplifies the linearity and grade coding, and assigns each risk-outcome pair a numeric code. 

```{r dataclean}
associations <- associations[!is.na(associations$Exposure),]

associations$Linearity <- sub(" .*", "", associations$Linearity)

associations$Linearity <- gsub("[–—−]", "-", 
                            gsub("[[:space:]]+", "", 
                              gsub("\u00A0", "", 
                                trimws(tolower(associations$Linearity)))))  

associations$Grade2 <- sub("\u00A0.*", "", associations$Grade2) #strange spacing in the column
associations$Grade2 <- sub(" .*", "", associations$Grade2)
associations$Grade2 <- substr(associations$Grade2, 1, nchar(associations$Grade2) - 1) #remove subscript

associations$code <- row.names(associations)

main_key <- associations[,c(13,2,3,4,5,10,11)] #Exposure, Outcome, Grade, Linearity
```

#Step 3 - Functions of the Lines

```{r linefunction}

linear_function <-  function(R,d,x) {
  exp((log(R)/d)*x)
}

linear_function_CI <- function(R, lower, upper, d, x) {
  estimate <- exp((log(R)/d)*x)
  lower_bound <- exp(log(lower)/d*x)
  upper_bound <- exp(log(upper)/d*x)
  cat("x =", x, "\n")
  cat("Estimate =", estimate, "\n")
  cat("Lower bound =", lower_bound, "\n")
  cat("Upper bound =", upper_bound, "\n")
}

non_linear_function <- function(x_coords, y_coords) {
  function(z) {
    idx <- findInterval(z, x_coords, rightmost.closed = TRUE)
    ifelse(idx == 0, y_coords[1], 
           ifelse(idx > length(y_coords), y_coords[length(y_coords)], y_coords[idx])) #last point is continuous
  }
} 

```

#Step 4 - Linear Data
All associations are included - either linear or summary estimates provided. 

```{r lineardata}

linear_assoc <- main_key

linear_assoc$`Dose/Metric` <- gsub("\\s+", "", linear_assoc$`Dose/Metric`) #removes spaces

linear_assoc$`Dose/Metric` <- sub("([a-zA-Zµ]+)[0-9]+$", "\\1", linear_assoc$`Dose/Metric`) #remove trailing digits (superscript)

linear_assoc <- linear_assoc %>%
  extract(`Dose/Metric`, into = c("Dose", "Metric"),
          regex = "([0-9\\.]+)([a-zA-Zµ]+)", remove = TRUE) #dose and metric separated

linear_assoc$Dose <- as.numeric(linear_assoc$Dose)

#note that association code 63 is 100 to 200g

colnames(linear_assoc)[6] <- "RR" 

linear_assoc <- linear_assoc %>%
  extract(RR, 
          into = c("Estimate", "Lower", "Upper"),
          regex = "([0-9\\.]+)\\s*\\[([0-9\\.]+)-([0-9\\.]+)\\]",
          convert = TRUE)
```

#Step 4B - Linear Functions
Creates a list of the linear function according to their code. Uncertainty is replicated in upper and lower bounds. 

```{r linearfunctions}
linear_functions <- list()

for(i in seq_len(nrow(linear_assoc))){
  code_p <- linear_assoc$code[i]
  rr_p <- linear_assoc$Estimate[i]
  dose_p <- linear_assoc$Dose[i]
  
  linear_functions[[paste0("function_",code_p)]] <- local({
    R <- rr_p
    d <- dose_p
    function(x) linear_function(R,d,x)
  })
}
```

```{r linearfunctions CI}
linear_functions_CI <- list()

for (i in seq_len(nrow(linear_assoc))) {
  code_p <- linear_assoc$code[i]
  rr_p <- linear_assoc$Estimate[i]
  lower_p <- linear_assoc$Lower[i]
  upper_p <- linear_assoc$Upper[i]
  dose_p <- linear_assoc$Dose[i]
  
  linear_functions_CI[[paste0("function_CI_", code_p)]] <- local({
    R <- rr_p
    lower <- lower_p
    upper <- upper_p
    d <- dose_p
    function(x) linear_function_CI(R, lower, upper, d, x)
  })
}
```

#Step 5 - Non-Linear Curve Data

```{r nonlinear data}
non_linear_assoc <- main_key[main_key$Linearity=="non-linear",]

non_linear_assoc <- non_linear_assoc %>%
  extract(`Dose/Metric`, into = c("Dose", "Metric"),
          regex = "([0-9\\.]+)([a-zA-Zµ]+)", remove = TRUE) #dose and metric separated
```

#Step 5B - Non-Linear Curves

```{r nonlinear curves}
csv_files <- list.files(path = file.path("data", "CSV coordinates"), pattern = "\\.csv$", full.names = TRUE)
csv_list <- lapply(csv_files, read.csv)
names(csv_list) <- tools::file_path_sans_ext(basename(csv_files))
```

#Step 5C - Non-Linear Functions

```{r nonlinear functions}
nonlin_functions <- lapply(csv_list, function(df) {
  x <- df[[1]]  
  y <- df[[2]]  
  non_linear_function(x, y)
})

new_names <- paste0("function_", sub("^code", "", names(csv_list)))
names(nonlin_functions) <- new_names
```

#Step 6 - Plots 

```{r linear}
output_dir <- file.path("plots")
dir.create(output_dir, showWarnings = FALSE)

linear_plots <- linear_assoc[linear_assoc$Linearity == "linear",]

for (i in seq_len(nrow(linear_plots))) {
  
  # Row-specific values
  code_p <- linear_plots$code[i]
  R <- linear_plots$Estimate[i]
  lower <- linear_plots$Lower[i]
  upper <- linear_plots$Upper[i]
  d <- linear_plots$Dose[i]
  
  unit <- linear_plots$Metric[i]
  if (is.na(unit) || unit == "") unit <- "unknown unit"
  
  exposure <- linear_plots$Exposure[i]
  outcome <- linear_plots$Outcome[i]  
  title <- paste(exposure, "and", outcome)
  x_label <- paste0("Intake ", unit, "/day")
  
  # X axis goes from zero to dose
  x_vals <- seq(0, d, length.out = 200)
  
  # Compute estimate, lower, upper
  estimate <- exp((log(R)/d) * x_vals)
  lower_bound <- exp((log(lower)/d) * x_vals)
  upper_bound <- exp((log(upper)/d) * x_vals)
  
  # Output files
  png_filename <- file.path(output_dir, paste0("plot_linear_", code_p, ".png"))
  png(png_filename, width = 800, height = 600)
  
  # Plot setup
  plot(x_vals, estimate, type = "n", 
       xlab = x_label, ylab = "Relative Risk", main = title,
       xlim = c(0, d), ylim = range(c(lower_bound, upper_bound, estimate)))
  
  # Uncertainty bands
  polygon(c(x_vals, rev(x_vals)), c(upper_bound, rev(lower_bound)),
          col = rgb(0, 0, 1, 0.2), border = NA)
  
  # Draw main line and points
  lines(x_vals, estimate, col = "blue", lwd = 2)
  points(x_vals, estimate, col = "blue", pch = 16, cex = 0.6)
  
  dev.off()
}
```

```{r nonlinear}
output_dir <- file.path("plots")
dir.create(output_dir)

for (i in seq_along(csv_list)) {
  csv_data <- csv_list[[i]]
  x_vals <- csv_data[[1]]
  y_vals <- csv_data[[2]]
  
  fn_name <- if (!is.null(names(csv_list))) names(csv_list)[i] else paste0("Function_", i)

  match_key <- sub("^code", "", fn_name)
  
  unit <- non_linear_assoc$Metric[i]
  if (is.na(unit) || unit == "") unit <- "unknown unit"
  
  exposure <- non_linear_assoc$Exposure[i]
  outcome <- non_linear_assoc$Outcome[i]  
  title <- paste(exposure, "and", outcome)
  x_label <- paste0("Intake ", unit, "/day")
  ord <- order(x_vals)
  x_vals <- x_vals[ord]
  y_vals <- y_vals[ord]

  #File name
  png_filename <- file.path(output_dir, paste0("plot_", fn_name, ".png"))
  
  #Plots
  png(png_filename, width = 600, height = 400)

    plot(NULL, xlim = c(min(x_vals) - 1, max(x_vals) + 1),
       ylim = range(y_vals) + c(-0.5, 0.5),
       xlab = x_label, ylab = "Relative Risk",
       main = title,  
       type = "n")

  for (j in 1:(length(x_vals) - 1)) {
    segments(x_vals[j], y_vals[j], x_vals[j + 1], y_vals[j], col = "blue", lwd = 2)
    points(x_vals[j], y_vals[j], pch = 16, col = "blue")
    points(x_vals[j + 1], y_vals[j], pch = 1, col = "blue")
  }

  segments(x_vals[length(x_vals)], y_vals[length(y_vals)],
           max(x_vals) + 2, y_vals[length(y_vals)], col = "blue", lwd = 2)
  points(x_vals[length(x_vals)], y_vals[length(y_vals)], pch = 16, col = "blue")

  dev.off()
}

```

#Step 7 - Use and Example

There are now 101 functions that can be used as well as plot outputs for visualisation of curves. Note that uncertainty is not accounted for here and the user will need to determine how they would like to estimate uncertainty in their given study. To use the functions, one simply needs to call it as seen below.

```{r example of function use}
#I would like to know the risk estimate for dementia based on fish intake (this is a linear association - code66)
linear_functions$function_66(10) #10g/day
linear_functions$function_66(50) #50g/day
linear_functions$function_66(100) #100g/day
linear_functions_CI$function_CI_66(10) #gives the estimate along with CI

#What about SSBs intake and Prostate Cancer incidence? (non-linear association - code18)
nonlin_functions$function_18(10)
nonlin_functions$function_18(50)
nonlin_functions$function_18(100)
```