---
title: "Dietary Risk Factor Dose-Response Curves V2.0"
author: "A. Jacob"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
```

#Step 1 - Import Data
This is an excel version of Table 2 from the manuscript located in the folder "data". 

```{r summary}
associations <- read_excel(file.path("data","Table2_all.xlsx"))
```

#Step 2 - Clean the Data
This step removes food category headers, simplifies the linearity and grade coding, and assigns each risk-outcome pair a numeric code. 

```{r dataclean}
associations <- associations[!is.na(associations$Exposure),]

associations$Linearity <- sub(" .*", "", associations$Linearity)

associations$Linearity <- gsub("[–—−]", "-", 
                            gsub("[[:space:]]+", "", 
                              gsub("\u00A0", "", 
                                trimws(tolower(associations$Linearity)))))  

associations$Grade2 <- sub("\u00A0.*", "", associations$Grade2) #strange spacing in the column
associations$Grade2 <- sub(" .*", "", associations$Grade2)
associations$Grade2 <- substr(associations$Grade2, 1, nchar(associations$Grade2) - 1) #remove subscript

associations$code <- row.names(associations)

main_key <- associations[,c(13,2,3,4,5,10,11)] #Exposure, Outcome, Grade, Linearity
```

#Step 3 - Functions of the Lines

```{r linefunction}

linear_function <-  function(R,lower,upper,d,x) {
  estimate <- exp((log(R)/d)*x)
  lower_bound <- exp(log(lower)/d*x)
  upper_bound <- exp(log(upper)/d*x)
}

non_linear_function <- function(x_coords, y_coords) {
  function(z) {
    idx <- findInterval(z, x_coords, rightmost.closed = TRUE)
    ifelse(idx == 0, y_coords[1], 
           ifelse(idx > length(y_coords), y_coords[length(y_coords)], y_coords[idx])) #last point is continuous
  }
} 

```

#Step 4 - Linear Data
All associations are included - either linear or summary estimates provided. 

```{r lineardata}

linear_assoc <- main_key

linear_assoc$`Dose/Metric` <- gsub("\\s+", "", linear_assoc$`Dose/Metric`) #removes spaces

linear_assoc$`Dose/Metric` <- sub("([a-zA-Zµ]+)[0-9]+$", "\\1", linear_assoc$`Dose/Metric`) #remove trailing digits (superscript)

linear_assoc <- linear_assoc %>%
  extract(`Dose/Metric`, into = c("Dose", "Metric"),
          regex = "([0-9\\.]+)([a-zA-Zµ]+)", remove = TRUE) #dose and metric separated

linear_assoc$Dose <- as.numeric(linear_assoc$Dose)

#note that association code 63 is 100 to 200g

colnames(linear_assoc)[6] <- "RR" 

linear_assoc <- linear_assoc %>%
  extract(RR, 
          into = c("Estimate", "Lower", "Upper"),
          regex = "([0-9\\.]+)\\s*\\[([0-9\\.]+)-([0-9\\.]+)\\]",
          convert = TRUE)
```

#Step 4B - Linear Functions
Creates a list of the linear function according to their code. Uncertainty is replicated in upper and lower bounds. 

```{r linearfunctions}
linear_functions <- list()

for(i in seq_len(nrow(linear_assoc))){
  code_p <- linear_assoc$code[i]
  rr_p <- linear_assoc$Estimate[i]
  lower_p <- linear_assoc$Lower[i]
  upper_p <- linear_assoc$Upper[i]
  dose_p <- linear_assoc$Dose[i]
  
  linear_functions[[paste0("function_",code_p)]] <- local({
    R <- rr_p
    lower <- lower_p
    upper <- upper_p
    d <- dose_p
    function(x) linear_function(R,lower,upper,d,x)
  })
}
```

#Step 5 - Non-Linear Curve Data

```{r nonlinear data}
non_linear_assoc <- main_key[main_key$Linearity=="non-linear",]
```

#Step 5B - Non-Linear Curves

```{r nonlinear curves}
csv_files <- list.files(path = file.path("data", "CSV coordinates"), pattern = "\\.csv$", full.names = TRUE)
csv_list <- lapply(csv_files, read.csv)
names(csv_list) <- tools::file_path_sans_ext(basename(csv_files))
```

#Step 5C - Non-Linear Functions

```{r nonlinear functions}
nonlin_functions <- lapply(csv_list, function(df) {
  x <- df[[1]]  
  y <- df[[2]]  
  non_linear_function(x, y)
})

new_names <- paste0("function_", sub("^code", "", names(csv_list)))
names(nonlin_functions) <- new_names
```

#Step 6 - Plots 

```{r linear}
output_dir <- file.path("plots")
dir.create(output_dir, showWarnings = FALSE)


for (i in seq_len(nrow(linear_curve))) {
  code_p <- linear_curve$code[i]
  R <- linear_curve$rr[i]
  d <- linear_curve$value[i]
  
  # Generate smooth sequence for 0–5*d
  x_vals <- seq(0, 5 * d, length.out = 200)
  y_vals <- linear_function(R, d, x_vals)
  
  # Metadata
  unit <- linear_assoc$unit[linear_assoc$code == code_p]
  if (length(unit) == 0) unit <- "unknown unit"
  
  x_label <- paste("Intake (", unit, ")", sep = "")
  title <- paste(linear_assoc$Exposure[linear_assoc$code == code_p],
                 "and",
                 linear_assoc$`Outcome `[linear_assoc$code == code_p])
  
  # Save PNG
  png_filename <- file.path(output_dir, paste0("plot_linear_", code_p, ".png"))
  png(png_filename, width = 600, height = 400)
  
  plot(x_vals, y_vals,
       type = "l", lwd = 2, col = "blue",
       xlab = x_label, ylab = "Relative Risk",
       main = title,
       xlim = c(0, 5 * d),
       ylim = range(y_vals))
  
  points(x_vals, y_vals, pch = 16, col = "blue", cex = 0.6)
  
  dev.off()
}
```

```{r nonlinear}
output_dir <- file.path("plots")
dir.create(output_dir)

for (i in seq_along(csv_list)) {
  csv_data <- csv_list[[i]]
  x_vals <- csv_data[[1]]
  y_vals <- csv_data[[2]]
  
  fn_name <- if (!is.null(names(csv_list))) names(csv_list)[i] else paste0("Function_", i)

  match_key <- sub("^code", "", fn_name)
  
  unit <- non_linear_assoc$`Dose/Units `[non_linear_assoc$code == match_key]
  if (length(unit) == 0) {
    unit <- "unknown unit"  # fallback if no match
  }

  x_label <- paste("Intake (", unit, ")", sep = "")
  
  title <- paste(non_linear_assoc$Exposure[non_linear_assoc$code == match_key]," and ",non_linear_assoc$`Outcome `[non_linear_assoc$code == match_key])
  
  ord <- order(x_vals)
  x_vals <- x_vals[ord]
  y_vals <- y_vals[ord]

  png_filename <- file.path(output_dir, paste0("plot_", fn_name, ".png"))
  png(png_filename, width = 600, height = 400)

    plot(NULL, xlim = c(min(x_vals) - 1, max(x_vals) + 1),
       ylim = range(y_vals) + c(-1, 1),
       xlab = x_label, ylab = "Relative Risk",
       main = title,  
       type = "n")

  for (j in 1:(length(x_vals) - 1)) {
    segments(x_vals[j], y_vals[j], x_vals[j + 1], y_vals[j], col = "blue", lwd = 2)
    points(x_vals[j], y_vals[j], pch = 16, col = "blue")
    points(x_vals[j + 1], y_vals[j], pch = 1, col = "blue")
  }

  segments(x_vals[length(x_vals)], y_vals[length(y_vals)],
           max(x_vals) + 2, y_vals[length(y_vals)], col = "blue", lwd = 2)
  points(x_vals[length(x_vals)], y_vals[length(y_vals)], pch = 16, col = "blue")

  dev.off()
}

```
The error is thrown due to code95 not being codable into a plot. This function does not follow regular axis intervals. 

#Step 7 - Use and Example

There are now 101 functions that can be used as well as plot outputs for visualisation of curves. Note that uncertainty is not accounted for here and the user will need to determine how they would like to estimate uncertainty in their given study. To use the functions, one simply needs to call it as seen below.

```{r example of function use}
#I would like to know the risk estimate for dementia based on fish intake (this is a linear association - code66)
linear_functions$function_66(10) #10g/day
linear_functions$function_66(50) #50g/day
linear_functions$function_66(100) #100g/day

#What about whole grain intake and CVD incidence? (non-linear association - code18)
nonlin_functions$function_18(10)
nonlin_functions$function_18(50)
nonlin_functions$function_18(100)
```